Follow-up Samples

### 2017-03-14 submission/manuscript_outline_20161024.tex
manuscript update

\textbf{Results:} There was a larger change in the bacterial community
in response to lesion removal for carcinoma versus adenoma cases
(P-value \textless{} 0.05). Yet no difference was found in the relative
abundance of any OTU before and after treatment for adenoma or carcinoma
groups (P-value \textgreater{} 0.05). A lesion model had an AUC range of
0.692 - 0.761 and follow up samples decreased in the positive
probability of lesion versus initial samples (P-value \textgreater{}
0.05); suggesting a movement towards a normal bacterial community. An
initial sample model had an AUC range of 0.657 - 0.796 and had a
decrease positive probability for the follow up samples to be an initial
sample (P-value \textless{} 0.05). The lesion model used a total of 54
variables while the initial sample model used a total of 70 variables. A
total of 32 OTUs were common to both models with the majority of these
classifying to commensal bacteria (e.g. \emph{Lachnospiraceae},
\emph{Ruminococcaceae}, \emph{Bacteroides}, and \emph{Ruminococcus}).

\textbf{Conclusions:} Our data supports the hypothesis that the
bacterial community changes after treatment. Individuals with carcinoma
have more drastic differences to the overall community then those with
adenoma. Commensal bacteria were crucial for accurate model
classification, suggesting that these bacteria may be important to
initial polyp formation and transition to carcinoma.

@@

Recent studies in humans have shown that both the bacterial community
and specific members within it correlate with CRC pathogenesis
{[}3,4{]}. Further, Dejea, et al. observed that bacterial communities
are altered between normal and tumor tissue {[}5{]}. Mouse models of CRC
have further demonstrated the importance of the microbiome, both on a
community {[}3,6{]} and species level {[}4{]}, for tumorgenesis.
Collectively, these studies provide a tantalizing link between our gut
bacteria and CRC and suggest that biomarkers using our microbes could be
developed. Indeed, builidng models using 16S rRNA gene sequencing along
with clinical tests such as Fecal Immunoglobulin Test (FIT) result in
good predictions of CRC {[}7,8{]}. Although these studies show how our
gut bacteria can impact CRC progression via a changed community or
invations by more inflammatory bacteria {[}9{]}. They provide very
little information as to whether these changed communities rebound after
successful treatment of lesion (adenoma or carcinoma).

In this study we tested the hypothesis that there are detectable changes
to the bacterial community between pre- (initial) and post- (follow up)
treatment of lesion. We analyzed changes in alpha and beta diversity as
well as the relative abundance of specific Operational Taxonomic Units
(OTUs). We then investigated several important genera from oral microbes
that have been suggested to be important in CRC pathogenesis {[}9{]}. We
next utilized Random Forest to build two models: The first was built to
classify lesion versus non-lesion (normal) while the second was built to
classify initial versus follow up samples. Subsequent observations on
how these models and OTUs within them performed before and after
treatment helped inform us as to whether initial and follow up samples
were changing and whether it was towards a normal community. We also
investigated the two models for similar important OTUs to identify the
crucial OTUs involved with both classifying lesion or normal and initial
versus follow up samples. Finally, since treatment also included
chemotherapy and radiation therapy along with lesion removal, we tested
if these additionaly treatments significantly lowered any of the metrics
examined versus those who received only lesion removal. This study will
provide evidence as to whether treatment can influence the community and
if the CRC microbiome, identified in previous studies, persists.

@@

\textbf{\emph{The Lesion Model:}} We tested whether follow up samples
were more normal than initial samples by building a model that
classified lesion versus normal. This model had an AUC range of 0.692 -
0.761 after 100 iterations of 20 repeated 10-fold corss validations. The
ROC curve for the final lesion model used was within the observed range
of the 100-different test set AUC iterations {[}Figure 2A{]}. There was
a total of 54 OTUs that were used in this model {[}Figure 2B{]}.
\emph{Lachnospiraceae} (Otu000015) resulted in the largest decrease in
MDA {[}FIgure 2B{]}. Heavily represented genera within the model
included OTUs from \emph{Lachnospiraceae}, \emph{Bacteroides},
\emph{Roseburia}, \emph{Blautia}, and \emph{Ruminococcus}.

If there was movement towards a more normal community after treatment of
lesion, we would expect to find a decrease in the positive probability
of the follow up sample to be a lesion. We observed no such decrease in
positive probability for either those with adenoma or carcinoma
{[}FIgure 2C \& 2D{]} (P-value \textgreater{} 0.05). We also observed a
significant difference between the predicted and actual calls (P-value
\textless{} 0.05). This model correctly classified the one individual
who still had carcinoma on follow up and this individual's positive
probability of lesion increased between their initial and follow up
sample {[}Figure 2C{]}.

\textbf{\emph{The Treatment Model:}} To test if there were differences
that could be used to classify initial and follow up treatment samples
we built a model to classify initial samples. This treatment model had
an AUC range of 0.657 to 0.796. after 100 iterations of 20 repeated
10-fold cross validations. The test set AUC range for this model
performed better than the training set AUCs. There was a marked decrease
in the ROC curve for the final model used when compared to the 100 test
set AUC iterations {[}Figure 3A{]}. There was a total of 70 OTUs that
were used for this model {[}Figure 3B{]}. The variable that resulted in
the largest MDA was \emph{Ruminococcaceae} (Otu000278) {[}Figure 3B{]}.
Similarly, to the lesion model heavily represented genera included OTUs
from \emph{Lachnospiraceae}, \emph{Bacteroides}, \emph{Roseburia},
\emph{Blautia}, and \emph{Ruminococcus}.

If there were changes between intial and follow up samples we would
expect the positive probability of being an initial sample to be
decreased in the follow up. The is what we observed for the treatment
model (P-value \textless{} 0.001). When we separated lesion into adenoma
and carcinoma there was a decrease in positive arobability for both the
carcinoma {[}Figure 3C{]} (P-value \textless{} 0.001) and adenoma group
{[}FIgure 3D{]} \textless{} 0.001). For this model, there was no
difference between the predicted and actual classifications (P-value
\textgreater{} 0.05).

\textbf{\emph{Common OTUs to both Models:}} We next wanted to know what
predictors within the lesion model were also in the initial sample
model. The main purpose was to identify which TOUs could be important in
lesion formation and were impacted by treatment. When we compared the
two different models with each other there were a total of 32 common
OTUs. Some of the most common taxonomic identifications belonged to
\emph{Lachnospiraceae}, \emph{Bacteroides}, \emph{Roseburia},
\emph{Blautia}, \emph{Anaerostipes}, \emph{Dorea}, and
\emph{Ruminococcus}. These along withthe vast majority of the OTUs that
were common between models had classifications to bacteria typically
thought of as commensal {[}Table S2{]}.

\textbf{\emph{Chemotherapy and Radiation Differences:}} After observing
these changes due to treatment we assessed whether chemotherapy or
radiation impacted the observed results. Only the treatment model had a
significant change in positive probability for those treated with
chemotherapy or radiation versus those who received neither {[}Figure 4A
\& Table S3{]} (P-value \textless{} 0.05). Using the 32 common OTUs we
tested how chemotherapy or radiation affects important members of the
community. FOr both chemotherapy or radiation, the directionaly of
change of these OTUs were similar {[}Figure 4B \& 4C{]}. After multiple
comparison correction, only \emph{Blautia} (Otu000006) remained
significantly different between those who did and did not receive
chemotherapy. This data suggests that follow up samples from those
treated with either chemotherapy or radiation may have had a larger
change from initial sample due to affects on the bacterial community.

@@

This study builds upon previous work from numerous labs that have
considered both how the bacterial community between those with and
without CRC differ and how it might be used as an early screening tool
{[}7,8,10--12{]}. Here we describe how the bacterial community changes
after treatment and which OTUs may be most important to lesion
classification. Interestingly, many of the most important OTUs for both
treatment and lesion classification had taxonomic identification for
resident gut microbes which could suggest that members within the
commensal community may be the first that change during pathogenesis of
disease. Further, these changes could be the first step in allowing more
inflammatory bacterium to gain a foothold within the colon {[}9{]}.

@@

Curiously, we observed that the typical CRC associated bacteria were not
predictive within our models. There are several reasons why this may
have occurred. First, is that they were not present in enough
individuals to be able to classify those with and without disease with a
high degree of accuracy. Second, is that our Random Forest models were
able to gather the same information from other OTUs. Third, it is also
possible that changes in commensal bacteria are the common differences
across adenoma and carcinoma. Finally, it is also possible that all
these explanations could have played a role. Our observations though,
would suggest that an individual's resident bacteria have a large role
to play in disease initiation and could change in a way that allows
predictive models to lower the positive probability of wither lesion or
initial sample before treatment {[}Figure 2 \& 3{]}. It should be noted
that our study does not argue against the importance of these CRC
associated bacteria in the pathogenesis of disease but rather that they
are not the main bacteria changing after treatment. It is possible that
these CRC associated bacteria are important in the transition from
adenoma to carcinoma and would be one explanation as to why in our data
we not only see high initial relative abundances in carcinoma and not
adenoma individuals but also large decreases in relative abundance in
some of those with carcinoma but not in those with adenoma after lesion
removal {[}Figure S2{]}.

>>>>

\textbf{Results:} There was a larger change in the bacterial community
in response to treatment for carcinoma versus adenoma cases (P-value
\textless{} 0.05). Yet no difference was found in the relative abundance
of any OTU before and after treatment for adenoma or carcinoma groups
(P-value \textgreater{} 0.05). A lesion model had an AUC range of 0.692
- 0.761 and follow up samples had no difference in the positive
probability of lesion versus initial samples (P-value \textgreater{}
0.05); suggesting that the lesion associated community persists after
treatment. A treatment model had an AUC range of 0.657 - 0.796 and had a
decrease positive probability for the follow up samples to be an initial
sample (P-value \textless{} 0.05); suggesting that there are members
within the community that repond to treatment. The lesion model used a
total of 54 variables while the initial sample model used a total of 70
variables. A total of 32 OTUs were common to both models with many
classifying to commensal bacteria (e.g. \emph{Lachnospiraceae},
\emph{Bacteroides}, \emph{Roseburia}, \emph{Blautia}, and
\emph{Ruminococcus}).

\textbf{Conclusions:} Our data partially supports the hypothesis that
the bacterial community changes after treatment. Individuals with
carcinoma have more drastic differences to the overall community then
those with adenoma. Commensal bacteria were crucial for accurate model
classification, suggesting that these bacteria may be important to
initial polyp formation and transition to carcinoma.

@@

Recent studies in humans have shown that both the bacterial community
and specific members within it correlate with CRC pathogenesis
{[}3,4{]}. Further, Dejea, et al. observed that bacterial communities
are altered between normal and tumor tissue {[}5{]}. Mouse models of CRC
have further demonstrated the importance of the microbiome, both on a
community {[}3,6{]} and species level {[}4{]}, for tumorgenesis.
Collectively, these studies provide a tantalizing link between our gut
bacteria and CRC and suggest that biomarkers using our microbes could be
developed. Indeed, builidng models using 16S rRNA gene sequencing along
with clinical tests such as Fecal Immunoglobulin Test (FIT) result in
good predictions of CRC {[}7,8{]}. Although these studies show how our
gut bacteria can impact CRC progression via a changed community or
invasion by more inflammatory bacteria {[}9{]}. They provide very little
information as to whether these changed communities rebound after
successful treatment of lesion (adenoma or carcinoma).

In this study, we tested the hypothesis that there are detectable
changes to the bacterial community between pre- (initial) and post-
(follow up) treatment of lesion. We analyzed changes in alpha and beta
diversity as well as the relative abundance of specific Operational
Taxonomic Units (OTUs). We then investigated several important genera
from oral microbes that have been suggested to be important in CRC
pathogenesis {[}9{]}. We next utilized Random Forest to build two
models: The first was built to classify lesion versus non-lesion
(normal) while the second was built to classify initial versus follow up
treatment samples. Subsequent observations on how these models and OTUs
within them performed before and after treatment helped inform us as to
whether they were changing after treatment and whether it was towards a
normal community. We also investigated the two models for similar
important OTUs to identify which were important for classifying both
lesion versus normal and initial versus follow up treatment samples.
Finally, since treatment also included chemotherapy and radiation
therapy along with lesion removal, we tested if these additionally
treatments significantly lowered any of the metrics examined versus
those who received only lesion removal. This study helps to provide
evidence as to whether treatment can influence the community and if the
CRC microbiome, identified in previous studies, persists.

@@

\textbf{\emph{The Lesion Model:}} We tested whether follow up samples
were more normal than initial samples by building a model based solely
on OTUs alone that classified lesion versus normal. This model had an
AUC range of 0.692 - 0.761 after 100 iterations of 20 repeated 10-fold
cross validations. The ROC curve for the final lesion model used was
within the observed range of the 100-different test set AUC iterations
{[}Figure 2A{]}. There was a total of 54 OTUs that were used in this
model {[}Figure 2B{]}. \emph{Lachnospiraceae} (Otu000015) resulted in
the largest decrease in MDA {[}Figure 2B{]}. Heavily represented genera
within the model included OTUs from \emph{Lachnospiraceae},
\emph{Bacteroides}, \emph{Roseburia}, \emph{Blautia}, and
\emph{Ruminococcus}.

If there was movement towards a more normal community after treatment of
lesion, we would expect to find a decrease in the positive probability
of the follow up sample to be a lesion. We observed no such decrease in
positive probability for either those with adenoma or carcinoma
{[}Figure 2C \& 2D{]} (P-value \textgreater{} 0.05). We also observed a
significant difference between the predicted and actual calls (P-value
\textless{} 0.05). However, this model correctly classified the one
individual who still had carcinoma on follow up and this individual's
positive probability of lesion increased between their initial and
follow up sample {[}Figure 2C{]}.

\textbf{\emph{The Treatment Model:}} To test if there were differences
that could be used to classify initial and follow up treatment samples
we built a model to classify initial samples. This treatment model had
an AUC range of 0.657 to 0.796. after 100 iterations of 20 repeated
10-fold cross validations. The test set AUC range for this model
performed better than the training set AUCs. There was a marked decrease
in the ROC curve for the final model used when compared to the 100 test
set AUC iterations {[}Figure 3A{]}. There was a total of 70 OTUs that
were used for this model {[}Figure 3B{]}. The variable that resulted in
the largest MDA was \emph{Ruminococcaceae} (Otu000278) {[}Figure 3B{]}.
Similar to the lesion model, heavily represented genera included OTUs
from \emph{Lachnospiraceae}, \emph{Bacteroides}, \emph{Roseburia},
\emph{Blautia}, and \emph{Ruminococcus}.

If there were changes between initial and follow up samples we would
expect the positive probability of being an initial sample to be
decreased in the follow up. The is what we observed for the treatment
model (P-value \textless{} 0.001). When we separated lesion into adenoma
and carcinoma there was a decrease in positive probability for both the
carcinoma {[}Figure 3C{]} (P-value \textless{} 0.001) and adenoma group
{[}Figure 3D{]} \textless{} 0.001). For this model, there was no
difference between the predicted and actual classifications (P-value
\textgreater{} 0.05).

\textbf{\emph{Common OTUs to both Models:}} We next wanted to know what
predictors within the lesion model were also in the initial sample
model. The main purpose was to identify which OTUs could be important in
lesion formation and were impacted by treatment. When we compared the
two different models with each other there were a total of 32 common
OTUs. Some of the most common taxonomic identifications belonged to
\emph{Lachnospiraceae}, \emph{Bacteroides}, \emph{Roseburia},
\emph{Blautia}, \emph{Anaerostipes}, \emph{Dorea}, and
\emph{Ruminococcus}. These along withthe vast majority of the OTUs that
were common between models had classifications to bacteria typically
thought of as commensal {[}Table S2{]}.

\textbf{\emph{Chemotherapy and Radiation Differences:}} After observing
these changes due to treatment we assessed whether chemotherapy or
radiation impacted the observed results. Only the treatment model had a
significant change in positive probability for those treated with
chemotherapy or radiation versus those who received neither {[}Figure 4A
\& Table S3{]} (P-value \textless{} 0.05). Additionally, there was no
difference between the change in positive probability for chemotherapy
and radiation therapy (Uncorrected P-value \textgreater{} 0.05). Using
the 32 common OTUs we tested how chemotherapy or radiation affects
important members of the community. For both chemotherapy and radiation,
the directionality of change of these OTUs were similar {[}Figure 4B \&
4C{]}. After multiple comparison correction, only \emph{Blautia}
(Otu000006) remained significantly different between those who did and
did not receive chemotherapy. This data suggests that follow up samples
from those treated with either chemotherapy or radiation may have had a
larger change from initial sample due to effects on the bacterial
community.

@@

This study builds upon previous work from numerous labs that have
considered both how the bacterial community between those with and
without CRC differ and how it might be used as an early screening tool
{[}7,8,10--12{]}. Here we describe how the bacterial community changes
after treatment and which OTUs may be most important to lesion
classification. Interestingly, many of the most important OTUs for both
treatment and lesion classification had taxonomic identification for
resident gut microbes. This suggest that members within the commensal
community may be the first that change during CRC pathogenesis. These
changes, in turn, could be the first step in allowing more inflammatory
bacteria to gain a foothold within the colon {[}9{]}.

@@

Curiously, we observed that the typical CRC associated bacteria were not
predictive within our models. There are several reasons why this may
have occurred. First, is that they were not present in enough
individuals to be able to classify those with and without disease with a
high degree of accuracy. Second, is that our Random Forest models were
able to gather the same information from other OTUs. Third, is that
changes in commensal bacteria are the common differences across adenoma
and carcinoma. Finally, it is also possible that all these explanations
could have played a role. Our observations though, would suggest that an
individual's resident bacteria have a large role to play in polyp
formation and could change in a way that allows predictive models to
lower the positive probability of lesion or initial sample before
treatment {[}Figure 2 \& 3{]}. It should be noted that our study does
not argue against the importance of these CRC associated bacteria in the
pathogenesis of disease but rather that they are not the main bacteria
changing after treatment. It is possible that these CRC associated
bacteria are important in the transition from adenoma to carcinoma and
would be one explanation as to why in our data we not only see high
initial relative abundances in carcinoma and not adenoma individuals but
also large decreases in relative abundance in some of those with
carcinoma but not in those with adenoma after treatment {[}Figure S2{]}.
